Characteristic polynomials of p-adic matrices

### 2017-02-03 charpoly/charpoly.tex
add \noindent

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \affaddr{\textsf{xavier.caruso@normalesup.org}}
\alignauthor David Roe\\
  \affaddr{University of Pittsburg}\\
  \affaddr{\textsf{roed@pitt.edu}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e de Limoges}\\
  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
}

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi_M$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

\begin{proof}
The equivalence of \eqref{pt:cyc} and \eqref{pt:1d} is standard; see
\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
that \eqref{pt:surj} and \eqref{pt:cyc} are equivalent.

For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
\end{proof}

We also have an analogue of Proposition \ref{prop:surjectivity} for integral matrices.

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
\end{proof}

Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ satisfies one of the conditions of Proposition \ref{prop:surjectivity}, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

\begin{proof}

@@

The relationship between precision and the images of lattices under $d\chi_M$ allows us to
apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
is the minimum possible.

\begin{cor}
Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^N)$.
Then the characteristic polynomial of $M$ has precision larger than $O(p^N)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

@@

\subsection{Stability under multiplication by $X$}

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$, and $V = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $V$ is stable under 
multiplication by $A$.
\end{prop}
\begin{proof}
Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $V$ is given by
the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
with its comatrix is the determinant, $(X - M) \cdot C = \chi(M)$ and thus
$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $P(X) \cdot V$, while the span of the entries
of the right hand side is contained within $V$ since $P(M) \in M_n(\OK)$.
\end{proof}

\begin{cor}
If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

We can also give a criterion on $M$ that allows an easy description of the image of flat
precision lattices under $d\chi(M)$.
\begin{prop}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then $d\chi_M\big(M_n(\OK)\big) = \OK[X] / \chi(M)$.
\end{prop}
\begin{proof}
It suffices to check that $d\chi_M$ is surjective modulo $\pi$.
Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.
\end{proof}

@@

In this section, we combine the computation of an Hessenberg form
of a matrix and the computation of the inverse through the SNF
over a complete discrete valuation field
to compute $\com (X-M)$ and $d \chi$
in a way such that if $M \in M_n(O_K),$
only division by invertible elements of $O_K$
occur.
%The computation of an Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

We begin with the study of the computation of
an approximate Hessenberg form.

@@

The second ingredient we need is the reciprocal polynomial.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $P \in K[X]$ be of degree $d \leq n.$ 
We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
whose coefficients are the reciprocal polynomials of the
coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
com(1-XM)^{rec}=&com(X-M), \\
(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
\end{eqnarray*}
\end{lem}

This lemma allows us to compute $com(1-XM)$ instead of $com(X-M).$
This has a remarkable advantage: the pivots during the computation of
the SNF of $com(1-XM)$ are units of $O_K[[X]],$ and are known
in advance to be on the diagonal. This leads to a very smooth
precision and complexity behaviour when the entry matrix lives in 
$M_n(O_K).$

@@

@@

\noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
by inverting $C_{1,1} \mod \chi(M).$

@@

{\bf Numerical experiments}

>>>>

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \affaddr{\textsf{xavier.caruso@normalesup.org}}
\alignauthor David Roe\\
  \affaddr{Pittsburg University}\\
  \affaddr{\textsf{roed@pitt.edu}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e de Limoges}\\
  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
}

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}
Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

\begin{proof}
We first show that the distinct generalized eigenvalues condition
guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
matrices with a single $1$ entry and all other entries $0$, it suffices to show
that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
are those on or below the diagonal within the Jordan blocks, and within the $i$th
block, each entry along the $j$th subdiagonal is
$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
As long as the $\lambda_i$ are distinct, these form a basis for $W$.

@@

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item the differential $d\chi(M)$ is surjective
\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
closure of $\bar K$ have all dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is stable by 
multiplication by the elements of $A$.
\end{prop}
\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
\end{proof}

\begin{cor}
If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

\begin{prop}
Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
\end{prop}
\begin{proof}
\todo{Finish proof}
\end{proof}

We can also give a criterion on $M$ that allows an easy description of the image of flat
precision lattices under $d\chi(M)$.
\begin{prop}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
\end{prop}
\begin{proof}
First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.
\end{proof}

@@

@@

In this section, we combine the computation of an Hessenberg form
of a matrix and the computation of the inverse through the SNF
over a complete discrete valuation field
to compute $\com (X-M)$ and $d \chi.$
%The computation of an Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

\todo{Describe algorithm (with matrices and endormophisms?). 
Remark about precision.}

@@

The second ingredient we need is the reciprocal polynomial.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $P \in K[X]$ be of degree $d \in \mathbb{N}_{>0}.$ 
We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
whose coefficients are the reciprocal polynomials of the
coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
com(1-XM)^{rec}=&com(X-M), \\
(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
\end{eqnarray*}
\end{lem}

This lemma allows us to compute $com(1-XM)$ instead of $com(X-M).$
This has a remarkable advantage: the pivots during the computation of
the SNF of $com(1-XM)$ are units of $O_K[[X]],$ and are known
in advance to be on the diagonal. This leads to a very smooth
precision and complexity behaviour.

@@

\todo{Add a small intro}

%Thanks to the previous results, we can now compute 
%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
%using the following algorithm.

@@

\noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
using Proposition \ref{prop:factor_com}.

@@

\noindent
{\bf Numerical experiments}
